Stage-Splitting a Modal Language

### 2014-07-25 paper/semantics.tex
Making an editing pass through 3.2.
My focus is eliminating some ambiguous wording in the exposition.  In
particular, I fear there is conflation of the idea of a lambda12
expression at stage 2), and the concept of the residual.

Secondly, I'd like to canonicalize the residual depiction in the
no-duplicate and duplicate cases.

The grammar and type system for \lang\ is given in
\ref{fig:grammar,fig:statics}. (Typing judgments and context variables are annotated with stages.)
Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.
%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.  
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
% We annotate typing judgments and context variables with stages;
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality\,\cite{cave14}.
The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
Although it is possible to implement $\pause$ from other \lang\ features, 
we provide it as a core primitive to simplify our examples.

@@

Although the code for {\tt fexp} looks very much like the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
%the decomposition of the exponent can be done at the stage \bbone\ without
%requiring disruption of the elegant functional structure of the code.}
We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.

A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as in the example above), this strategy conflicts with the precept of staged execution: that all stage \bbone\ code be evaluated before the evaluation of stage \bbtwo\ code. 
Accordingly, a suitably staged dynamic semantics for \lang\ must evaluate
\emph{all} the stage \bbone\ subexpressions of a term before its stage
\bbtwo\ subexpressions.

@@

To understand the difficulty of staged evaluation, consider the following
example:
\begin{lstlisting}
#2 (next {1+2}, 3+4)
\end{lstlisting}
This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. A conventional
call-by-value evaluation strategy demands that we evaluate both components of
the pair to values before we project from the pair. However, at stage \bbone\ we
are not allowed to evaluate under the \verb|next|; on the other hand, 
\verb|next {1+2}| is not a value, in the sense that it requires additional
computation steps.

Intuitively, the solution is to say that \verb|next {1+2}| is a value \emph{at
stage \bbone}, even though it requires additional evaluation at stage \bbtwo.
Therefore, we evaluate the pair to
\begin{lstlisting}
#2 (next {1+2}, 7)
\end{lstlisting}
and get \verb|7| as the answer.

To complicate matters, consider a situation where we must substitute such a
incompletely-evaluated expression. The following stage-\bbtwo\ term has type
$\rmint$:
\begin{lstlisting} 
prev{
  let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
Stage-\bbone\ evaluation will not fully reduce this term, because the answer
depends on the value of \verb|1+2|, but we cannot reduce this to \verb|3| until
stage \bbtwo.

If we simply treat \verb|(next {1+2}, 7)| as a value during stage \bbone, then
we will substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression. Thus, the result of stage \bbone\ computation is
\begin{lstlisting} 
prev{
  next{ prev{next {1+2}} * prev{next {1+2}} * 7 }
}
\end{lstlisting}
When we evaluate this at stage \bbtwo, we will compute \verb|1+2| twice. Of
course, this is not particularly expensive, but in general, the duplicated
computation could be anything, even a recursive call to the function!

To avoid duplicating computations, we take a different approach.  Instead of
duplicating the contents of the first $\next$ expression, we bind them to some
variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
This produces:
\begin{lstlisting} 
let y = 1+2 in y * y * 7
\end{lstlisting}

Achieving this behavior mechanically requires us to resolve a contradiction:
we must substitute for \texttt{x} at stage \bbone, but we cannot evaluate inside the $\next$ block bound to it. 
Our solution is to replace the contents of the $\next$ with a new variable and create an explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s old contents.  
This substitution then floats up to the top of the containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->1+2]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
As a convention, we render the new variable with a %stylish and fashionable
hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
\begin{lstlisting} 
prev {
[y|->1+2]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them. 
As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
\begin{lstlisting} 
prev {
[yhat|->1+2]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->1+2]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = 1+2 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}

>>>>

The grammar and type system for \lang\ is given in
\ref{fig:grammar,fig:statics}. (Typing judgments and context variables are annotated with stages.)
Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.
%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.  
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
% We annotate typing judgments and context variables with stages;
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality\,\cite{cave14}.
The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
It is possible to implement $\pause$ from other \lang\ features, but
we provide it as a core primitive to simplify our examples.

@@

Although the code for {\tt fexp} appears very similar to the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
%the decomposition of the exponent can be done at the stage \bbone\ without
%requiring disruption of the elegant functional structure of the code.}
%
% NOTE(kayvonf): removing the following comments about binding time analysis from the text for now to improve clarity.  They will certainly be addressed in some prior section in the paper.
%
%We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
%
A key attribute of this example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as the recurive call to {\tt fexp} above), this strategy conflicts with the precept of staged execution: that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code. 
Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
\emph{all} the stage \bbone\ subexpressions of a term before its stage
\bbtwo\ subexpressions.

@@

To gain intuition about the challenges of ensuring staged evaluation of \lang, consider the following
example:
\begin{lstlisting}
#2 (next {1+2}, 3+4)
\end{lstlisting}
This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. A conventional
call-by-value evaluation strategy demands that we evaluate both components of
the pair to values before we project from the pair. The problem is that
while \verb|next {1+2}| is not a value (in the sense that additional
stage-\bbtwo\ computation steps are necessary to produce the result \verb|3|), evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
Intuitively, the solution is to designate \verb|next {1+2}| as a value \emph{in
stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
Therefore, we evaluate the pair to
\begin{lstlisting}
#2 (next {1+2}, 7)
\end{lstlisting}
then obtain \verb|7| as the result of projection.

Now consider a more complex example where 
stage-\bbone\ evaluation must substitute such a incompletely-evaluated
expression. The following stage-\bbtwo\ term has type $\rmint$:
\begin{lstlisting} 
prev{
  let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
Again, stage-\bbone\ evaluation will not fully reduce this term, because the answer
depends on the value of \verb|1+2|, which is not reduced to \verb|3| until
stage \bbtwo.

If we simply treat \verb|(next {1+2}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the result of stage \bbone\ computation is
\begin{lstlisting} 
prev{
  next{ prev{next {1+2}} * prev{next {1+2}} * 7 }
}
\end{lstlisting}
Stage-\bbtwo\ evaluation of this expression will compute \verb|1+2| twice. Of
course, this is not particularly expensive, but in general, the duplicated
computation could be anything, even a recursive call to the function! To avoid duplicating
computations, we take a different approach.  Instead of
duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
a variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
This produces:
\begin{lstlisting} 
let y = 1+2 in y * y * 7
\end{lstlisting}

Achieving this behavior mechanically requires us to resolve a contradiction:
we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block bound to it. 
Our solution is to replace the contents of the $\next$ with a new variable and create an explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s old contents.  
This substitution then floats up to the top of the containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->1+2]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
As a convention, we render the new variable with a %stylish and fashionable
hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
\begin{lstlisting} 
prev {
[y|->1+2]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them. 
As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
\begin{lstlisting} 
prev {
[yhat|->1+2]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->1+2]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = 1+2 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}
